---
title: 'Creating Engaging Plots in ggplot2: Exercises'
author: "Angela Li"
date: "2023-02-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Gapminder data

```{r wd, warning=F}
# Load any packages needed
library(tidyverse)

# Set working directory (in fact, unnecssary when using R projects!) and read in data
gapminder <- read.csv("gapminder.csv")
```

We'll start with a few commands to explore our data. This data is from [Gapminder](https://www.gapminder.org/data/documentation/), and has been edited for use in [open-source Data Carpentry materials](https://datacarpentry.org/r-intro-geospatial/). (If you want to know more, there's a [2006 TED talk](https://www.ted.com/talks/hans_rosling_the_best_stats_you_ve_ever_seen) with this data.) Our research question is how life expectancy at birth compares across countries over time.
 
The codebook from the survey 
can be found online [here](https://addhealth.cpc.unc.edu/documentation/codebook-explorer/#/variable_collection/1573).

Columns in the dataset include:

- `country`
- `continent`
- `year`
- `lifeExp` or life expectancy
- `pop` or population
- `gdpPercap` or GDP per capita

```{r explore-data}
# Take a look at the data
head(gapminder)

# Find unique countries represented in the data
unique(gapminder$country)

# Filter to just one county for now
us <- filter(gapminder, country == "United States")
```

### Your Turn

Let's learn about this data! Answer the following:

1. What's the year span represented in the country you chose?
2. What does each row mean? Is this "long" or "wide" data?

### Your Turn - Bonus

Debugging is a large part of learning to code. Fix the following pieces of 
incorrect code! (Note - change eval = T so that the chunk runs when knitting the doc)

```{r debug, eval=F}
# Fix the following pieces of incorrect code!
summary(gapmider)

read.cvs("gapminder.csv")

filter(gapminder, county = "United States")

ggplot(gapminder, aes(x = age)) %>%
  geom_histogram()
```

# Visualizing our data with ggplot2

We'll start by looking at the trend in life expectancy for one country in our dataset before moving to multiple countries.

```{r viz-example}
# We start by specifying our data
ggplot(data = us) 

# We then map our *aes*thetics to axes
ggplot(data = us, aes(x = year, y = lifeExp))

# Then add the **geom**etry (or geometries) - note the plus sign!
ggplot(data = us, aes(x = year, y = lifeExp)) +
  geom_point()

ggplot(data = us, aes(x = year, y = lifeExp)) +
  geom_point() +
  geom_line()

# Add informative labels (and maybe a subtitle or caption)
ggplot(data = us, aes(x = year, y = lifeExp)) +
  geom_point() +
  geom_line() +
  labs(x = "Year",
       y = "Life expectancy",
       title = "Life expectancy in the US",
       subtitle = "Increased between 1952 and 2007",
       caption = "Source: Gapminder.org")

# Adjust the theme
ggplot(data = us, aes(x = year, y = lifeExp)) +
  geom_point() +
  geom_line() +
  labs(x = "Year",
       y = "Life expectancy",
       title = "Life expectancy in the US",
       subtitle = "Increased between 1952 and 2007",
       caption = "Source: Gapminder.org") +
  theme_bw()

# Adjust limits
ggplot(data = us, aes(x = year, y = lifeExp)) +
  geom_point() +
  geom_line() +
  labs(x = "Year",
       y = "Life expectancy",
       title = "Life expectancy in the US",
       subtitle = "Increased between 1952 and 2007",
       caption = "Source: Gapminder.org") +
  theme_bw() +
  ylim(0, 80)
```

### Your Turn

Visualize the change in **population** for your country of interest using ggplot2. 

```{r viz-prac}
ggplot(data = us, aes(x = year, y = pop)) +
  geom_point() +
  geom_line() +
  labs(x = "",
       y = "",
       title = "Population in the US",
       subtitle = "Steady increase between 1952 and 2007",
       caption = "Source: Gapminder.org") +
  theme_bw()
```

Try making a bar chart instead of a line graph of population (hint: use `geom_col`). Is this the best visualization for this data? Why not?

```{r viz-prac-2}
ggplot(data = us, aes(x = year, y = pop)) +
  geom_col() +
  labs(x = "",
       y = "",
       title = "Population in the US",
       subtitle = "Steady increase between 1952 and 2007",
       caption = "Source: Gapminder.org") +
  theme_bw()
```

## Plotting multiple countries

We want to compare multiple countries, so let's include 5 from multiple continents in our dataset.

```{r}
gap5 <- gapminder %>%
  filter(country %in% c("Sierra Leone", "United States", "Italy", "Nigeria", "India"))
```

Then, visualize the subset of the data:

```{r}
ggplot(data = gap5, aes(x = year, y = lifeExp, color = country)) +
  geom_smooth() +
  geom_point(alpha = 0.5) + # alpha makes points transparent
  labs(x = "",
       y = "Life expectancy at birth (years)",
       title = "Life expectancy over time",
       subtitle = "While generally increasing, HIV/AIDS impacted outcomes",
       caption = "Source: Gapminder.org") +
  theme_minimal() +
  theme(legend.title = element_blank()) + 
  guides(color=guide_legend(override.aes=list(fill=NA)))
```

### Your Turn

Create a plot for **population** over time for 5 countries in the dataset using ggplot2.

```{r pop-viz}

```

### Your Turn - Bonus

You can also create small multiples of graphs, where each country is represented in its own graph.

Poke around the [ggplot2 book on facetting](https://ggplot2-book.org/getting-started.html#qplot-faceting) and see if you can figure out how to make "small multiples" for the trend in each country.

```{r}
ggplot(data = gap5, aes(x = year, y = lifeExp)) +
  geom_smooth() +
  labs(x = "",
       y = "Life expectancy at birth (years)",
       title = "Life expectancy over time",
       caption = "Source: Gapminder.org") +
  theme(legend.title = element_blank()) + 
  theme_bw() +
  guides(color=guide_legend(override.aes=list(fill=NA))) + 
  facet_wrap(~country)
```

## Bar Charts

Now, we will subset to a given year and compare an outcome of our choice in 5 countries.

```{r}
gap5 %>%
  ggplot(aes(x = country, y = gdpPercap)) +
  geom_col()
```

### Your Turn

Improve on the plot we created to make it more engaging and informative to the viewer. If you want, try the `colors()` function in the R console to see the color names available. The [R Graph Gallery](https://r-graph-gallery.com/42-colors-names.html) also provides guidance.

```{r pop-viz2}
options(scipen = 999)

gap5 %>%
  ggplot(aes(x = country, y = gdpPercap)) +
  geom_col(fill = "darkorange3") + # Tr
  labs(x = "",
       y = "",
       title = "GDP Per Capita by Country",
       subtitle = "US and Italy have the highest GDP per capita in 2007",
       caption = "Source: Gapminder.org") +
  theme_bw()
```

### Your Turn - Bonus

The figure defaults in R don't order the bars in a meaningful way. There are two tips that will help with this:

1. Flipping the x and y axes: `+ coord_flip()`
2. Ordering bars in a meaningful way, ie by the levels, not alphabetically `x = fct_reorder(<things to be reordered>, <thing to reorder it by>)`

Try implementing this to reorder the bars!

```{r}
gap5 %>%
  ggplot(aes(x = country, y = gdpPercap)) +
  geom_col() +
  coord_flip() +
  theme_bw()
```

```{r}
gap5 %>%
  ggplot(aes(x = fct_reorder(country, gdpPercap), y = gdpPercap)) +
  geom_col() +
  coord_flip()
```

There's much more to ggplot2, but for that, I direct you to some much better guides that have been written...

### Your Turn: Explore Open Source Resources

Take 5 minutes and look through the resources listed below. Which one of these look the most useful for your work?

- [Posit Cloud Primers - Visualize Data](https://posit.cloud/learn/primers/3): Interactive online tutorials to learn data viz in R, from the makers of RStudio
- [R Graphics Cookbook, 2nd edition](https://r-graphics.org/): Gives you recipes for creating various graphs in R - https://r-graphics.org/
- [ggplot2 book](https://ggplot2-book.org/): Nitty gritty on how ggplot2 all works, from the creator of ggplot2
- [Data Visualization: A Practical Introduction]: Online book by Soc Prof Kieran Healy walking you through exercises in R
- [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/): Less specific to ggplot2, more on best practices to learn data viz
- [Data Visualization course by Andrew Heiss](https://datavizs21.classes.andrewheiss.com/): Designed as an entire course in R for policy school students

### Your Turn - Bonus

Schedule in an hour on your calendar (possibly over break) to explore that resource. 
