---
title: 'Creating Engaging Plots in ggplot2: Exercises'
author: "Your Name"
date: "2023-02-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Gapminder data

```{r wd, warning=F}
# Load any packages needed


# Set working directory (in fact, unnecssary when using R projects!) and read in data

```

We'll start with a few commands to explore our data. This data is from [Gapminder](https://www.gapminder.org/data/documentation/), and has been edited for use in [open-source Data Carpentry materials](https://datacarpentry.org/r-intro-geospatial/). (If you want to know more, there's a [2006 TED talk](https://www.ted.com/talks/hans_rosling_the_best_stats_you_ve_ever_seen) with this data.) Our research question is how life expectancy at birth compares across countries over time.

The codebook from the survey 
can be found online [here](https://addhealth.cpc.unc.edu/documentation/codebook-explorer/#/variable_collection/1573).

Columns in the dataset include:

- `country`
- `continent`
- `year`
- `lifeExp` or life expectancy
- `pop` or population
- `gdpPercap` or GDP per capita

```{r explore-data}
# Take a look at the data

# Find unique countries represented in the data

# Filter to just one county for now

```

### Your Turn

Let's learn about this data! Answer the following:

1. What's the year span represented in the country you chose?
2. What does each row mean? Is this "long" or "wide" data?

### Your Turn - Bonus

Debugging is a large part of learning to code. Fix the following pieces of 
incorrect code! (Note - change eval = T so that the chunk runs when knitting the doc)

```{r debug, eval=F}
# Fix the following pieces of incorrect code!
summary(gapmider)

read.cvs("gapminder.csv")

filter(gapminder, county = "United States")

ggplot(gapminder, aes(x = age)) %>%
  geom_histogram()
```

# Visualizing our data with ggplot2

We'll start by looking at the trend in life expectancy for one country in our dataset before moving to multiple countries.

```{r viz-example}
# We start by specifying our data

# We then map our *aes*thetics to axes

# Then add the **geom**etry (or geometries) - note the plus sign!

# Add informative labels (and maybe a subtitle or caption)

# Adjust the theme

# Adjust limits

```

### Your Turn

Visualize the change in **population** for your country of interest using ggplot2. 

```{r viz-prac}

```

Try making a bar chart instead of a line graph of population (hint: use `geom_col`). Is this the best visualization for this data?

```{r viz-prac-2}

```

## Plotting multiple countries

We want to compare multiple countries, so let's include 5 from multiple continents in our dataset.

```{r}

```

Then, visualize the subset of the data:

```{r}

```

### Your Turn

Create a plot for **population** over time for 5 countries in the dataset using ggplot2.

```{r pop-viz}

```

### Your Turn - Bonus

You can also create small multiples of graphs, where each country is represented in its own graph.

Poke around the [ggplot2 book on facetting](https://ggplot2-book.org/getting-started.html#qplot-faceting) and see if you can figure out how to make "small multiples" for the trend in each country.

```{r}

```

## Bar Charts

Now, we will subset to a given year and compare an outcome of our choice in 5 countries.

```{r}

```

### Your Turn

Improve on the plot we created to make it more engaging and informative to the viewer. 

```{r pop-viz}

```

### Your Turn - Bonus

The figure defaults in R don't order the bars in a meaningful way. There are two tips that will help with this:

1. Flipping the x and y axes: `+ coord_flip()`
2. Ordering bars in a meaningful way, ie by the levels, not alphabetically `x = fct_reorder(<things to be reordered>, <thing to reorder it by>)`

Try implementing this to reorder the bars!

```{r}

```

```{r}

```

There's much more to ggplot2, but for that, I direct you to some much better guides that have been written...

### Your Turn: Explore Open Source Resources

Take 5 minutes and look through the resources listed below. Which one of these look the most useful for your work?

- [**Posit Cloud Primers**](https://posit.cloud/learn/primers/3): Interactive online tutorials to learn data viz in R, from the makers of RStudio
- [**R Graphics Cookbook, 2nd edition**](https://r-graphics.org/): Gives you recipes for creating various graphs in R - https://r-graphics.org/
- [**ggplot2 book**](https://ggplot2-book.org/): Nitty gritty on how ggplot2 all works, from the creator of ggplot2
- [**Data Visualization: A Practical Introduction**](https://socviz.co/): Online book by Soc Prof Kieran Healy walking you through exercises in R
- [**Fundamentals of Data Visualization**](https://clauswilke.com/dataviz/): Less specific to ggplot2, more on best practices to learn data viz
- [**Data Visualization course by Andrew Heiss**](https://datavizs21.classes.andrewheiss.com/): Designed as an entire course in R for policy school students
- [**Tidy Tuesday R Community**](https://www.tidytuesday.com/): A community of folks learning R!

### Your Turn - Bonus

Schedule in an hour on your calendar (possibly over break) to explore that resource. 
