---
title: "Project 2"
author: "Your Name"
date: "2023-02-05"
output:
  html_document: default
---

```{r, echo=F, eval=F, message=F,warning=F}
#set working directory
setwd("~/Documents/workshop-1")

# you will only need to install packages once 
install.packages("tidyverse")
install.packages("kableExtra")
install.packages("stargazer")
install.packages("knitr")
install.packages("rdrobust")
```

```{r, echo=F, eval=T, message=F,warning=F}
# you will do this whenever 
# you need to load in and use these packages
library(tidyverse)
library(kableExtra)
library(stargazer)
library(knitr)
library(rdrobust)
```

## Overview

This document provides high-level information on whether the minimum legal drinking age (MLDA) makes any difference in the likelihood of consuming alcohol. It includes:

-   Code to read in and clean data

-   Summary Statistics

-   Regression output

## Data and Summary Statistics

The data come from a nationally representative survey, and the data come in the structure below:

```{r echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
# read in and inspect MLDA data 
mlda <- read_csv("mlda.csv")
mlda
```

```{r echo=F, eval=F, message=F,warning=F}
View(mlda)
```

## Some quick data exploration

Below is code and output for the share of the population that is a student and the share of a population that drinks alcohol.

```{r echo=FALSE, eval=FALSE, warning=FALSE, message=FALSE}
mean(mlda$student)
mean(mlda$drinks_alcohol)
```

-   The share who drink alcohol is `r round(mean(mlda$drinks_alcohol),2)` and the share who are students is `r round(mean(mlda$student),2)`.

## Transform and explore

Below is some code to generate summary statistics.

```{r echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
# add a new variable for age in years and a heavy drinker variable if they drink more than 75% of days
mlda <- mutate(mlda, 
               age_years = 21 + days_21/365, 
               heavy_drinker = if_else(perc_days_drink > 75, 1, 0))

# filter to just underage drinkers 
underage <- filter(mlda, days_21 < 0 & drinks_alcohol == 1)

# demographic characteristics of underage drinkers 
summ_stats_underage <- summarise(underage, 
          min_age = min(age_years),
          mean_age = mean(age_years), 
          median_age = median(age_years),
          mean_heavy_drinker = mean(heavy_drinker), 
          mean_white = mean(white), 
          mean_black = mean(black), 
          mean_hispanic = mean(hispanic))

summ_stats_underage

```

## The Pipe

-   The pipe operate `%>%` takes output from on function and 'pipes' it into another function



```{r, eval=F}
x <- paste("A", "String")
print(x)

print(paste("A", "String"))

paste("A", "String") %>% 
  print(.)
```

## Use the pipe operator to combine all of these into one 'dplyr chain'

-   The period denotes where the previous output should be the argument in the new function



```{r}
summ_stats <- mutate(mlda, 
                     age_years = 21 + days_21/365, 
                     heavy_drinker = if_else(perc_days_drink > 75, 1, 0)) %>% 
  filter(., days_21 < 0 & drinks_alcohol == 1) %>%  
  summarise(.,min_age = min(age_years),
          mean_age = mean(age_years), 
          median_age = median(age_years),
          mean_perc_days_drink = mean(perc_days_drink), 
          median_perc_days_drink = median(perc_days_drink)) 

summ_stats
```

## More Summary Statistics

Summary statistics for heavy drinkers vs non-heavy drinkers.

```{r, echo=F, eval=T, message = F, warning=F}
# now let's look at summary stats of heavy drinkers vs non heavy drinkers: 

# demographic characteristics of heavy drinkers vs non-heavy drinkers
summ_stats_heavy <- mlda %>% 
  filter(drinks_alcohol ==1) %>%
  group_by(heavy_drinker) %>%
  summarise(min_age = min(age_years),
          mean_age = mean(age_years), 
          median_age = median(age_years),
          mean_student = mean(student),
          n = n())
summ_stats_heavy
```

## Try yourself

-   Create a variable that indicates if someone is a student that drinks (they are a student and they drink)
-   Use the skeleton code below to calculate the share of student drinkers that are male, white, black, hispanic



```{r, eval=T, echo=F}
summ_stats2 <- mutate(mlda, 
                      student_drinker = if_else() %>%
  group_by() %>%  
  summarise() 
  
summ_stats2
```

## Export as a table

-   use kable to export as a table
-   can make aesthetic changes to the table, add labels, etc.

```{r, eval=T,echo=F, warning=F,message=F}
library(kableExtra)

kable(summ_stats_underage,
      digits=2, caption="Drinking Patterns among Underage Drinkers", 
      col.names = c("Min Age", "Mean Age", "Med Age", "Heavy Drinker", 
                    "White", "Black", "Hispanic")) 
```

## Visual: age distribution of drinkers vs non-drinkers?

```{r, eval=T, echo=F, fig.height=2.6,fig.width=4}
viz <- mlda %>% 
  mutate(drinks_alcohol_string = if_else(drinks_alcohol == 1, "Drinks", "Does Not Drink")) %>%
  ggplot(aes(x=drinks_alcohol_string, y=age_years)) +
  geom_boxplot()
viz
  
```

## Run a Model

-   Use a linear probability model to predict the effect of age in years on the likelihood of drinking alcohol
-   we can use lm()

## Model our data

```{r, echo=T, eval=TRUE}
# lpm of years in age on likelihood of drinking alcohol
lpm1 <- lm(formula = drinks_alcohol ~ age_years, data = mlda )
```

```{r, echo=T, eval=F}
summary(lpm1)
```

## Exercise

-   Now, add an age squared and age cubed variable to your dataset and rerun the model

```{r,eval=T,echo=F}
mlda2 <- mlda %>% 
  mutate(age_sq = , age_cu = ) 

lpm2 <- lm(formula = , data = )
```

```{r,eval=T,echo=F}

lpm3 <- lm(formula =, data=)
```

-   inspect your results

```{r,eval=F,echo=F}

summary(lpm2)
summary(lpm3)

```

## Use Stargazer to output your results

Below is the output from several statistical models. As you can see, it appears that the minimum legal drinking age seems to have an effect on the likelihood of consuming alcohol.

```{r, echo=F,eval=F,message=F,warning=F,results='asis'}
library(stargazer)
stargazer(lpm1, lpm2, type = "html", 
          dep.var.labels = "Probabilty of Drinking", header=F)
```

```{r, echo=F,eval=T,message=F,warning=F,results='asis'}
library(stargazer)

stargazer(lpm1, lpm2, 
          type = "html", 
          dep.var.labels = "Probabilty of Drinking", 
          header=F)
```

## Output model results

```{r, echo=F,eval=T,message=F,warning=F,results='asis'}
library(stargazer)
stargazer(lpm1, lpm2, lpm3, type = "html", dep.var.labels = "Probabilty of Drinking with Controls", header=F)
```

## Visualize the discontinuity

```{r, warning=F,message=F, echo=F,fig.height=2.6,fig.width=4}
library(rdrobust)
rdplot(mlda$drinks_alcohol,mlda$age_years,c=21)
```

## Conclusion and Takeaways

-   It seems that the minimum legal drinking age has some effect
